The Evolution of Phenotypically Invariant Genetic Networks

### 2021-08-10 review-responses2.tex

\reply{
This is correct. As the model is currently described, we show that since different mutationally connected systems are not usually reproductively compatible, populations in different parts of system space will produce unfit hybrids. We rely on results from quantitative genetics to estimate the rate at which systems drift through the neutral space and thus to estimate the effects of gene flow. Without further modeling we do not yet know how a quantitative genetics model incorporating system theory would differ, if it all, from these classic results. We are planning to submit a paper where we further develop the removed quantitative genetics model and may be able to address this interesting question there.

We also briefly mention that gene flow will likely complicate matters in the discussion \revref.
}

\begin{point}{Orr's Snowball}
The authors state that rapid emergence of incompatibility is possible in their model. How would the accumulation of incompatibility here compare with Orr’s ``snownball'' model, and how is this expected to depend on the complexity of the underlying networks?
\end{point}

\reply{
    Good question:
    on face value, the similarity to the snowball depends entirely on the relationship between
    ``distance to optimum'' and fitness: since distance to optimum increases linearly in F2s,
    if this relation is quadratic then our model will also show a quadratic increase in incompatibility.
    However, this seems somewhat arbitrary.
    It is more interesting to consider if our model displays the same underlying behavior that leads to the snowball,
    in which each pair of lineage-specific substitutions possibly contributes to incompatibility.
    This may be an empirical question (that could be examined in simulation),
    but at the level of analysis in our paper it seems the answer is ``no'':
    Consider two lineages moving away from each other along the neutral manifold.
    Each substitution moves them further apart, but the rate of increase of the distance between them
    is linear, and each subsequent substitution does not increase their rate of separation.
    We considered adding this point to the paper,
    but decided that it is sufficiently confusing that it would not add much.
}

@@

\reply{This is an important point. We have added references to the experimental speciation literature \revref. 
}

\begin{point}{Additive quantitative trait models}
%2. Additive quantitative trait models
%
The previous version of this manuscript contained a standard quantitative trait model. I think it was a good idea to remove this material, but I still think that this modelling approach (including Fisher’s geometric model) needs to be discussed a little more here. In the introduction, for example, it seems important to state clearly that ``kryptotypes'' (many-to-one genotype-to-phenotype maps) are also a feature of these standard models, and that previous authors have shown how phenotypically identical populations can evolve RI, in some cases via drift alone (e.g., Mani and Clarke 1990; Rosas et al. 2010; Fraisse et al. 2016 and others).

Martin 2014 argued that the results of these models and ``systems-biology'' type models are likely to approximate one another under some conditions, and that could be mentioned.
\end{point}

\reply{
    We agree that this is an important thing to discuss.
    We think that this is fairly well covered in the Introduction (paragraph at \revref),
    and cite \citet{rosas2010cryptic} and \citet{fraisse2016genetics} in that section --
    we have tried to make the connection to previous literature complete without being overwhelming.
    We have also added a reference to \citet{mani1990mutational} at \llname{citemani},
    and added further reference to Fisher's geometric model in the Discussion \llname{quantnote}.
}

@@

\begin{point}{RI due to kryptotypes not involving drift}
%3. RI due to kryptotypes not involving drift
%
  The section beginning on \hyperlink{rev1:4}{line 419} seems quite important to me, but it seems to downplay an important point. Populations might diverge genetically, or in their kryptotypes, but not in their phenotypes, if they experience fluctuating selection. This goes back the Wright-Fisher debates, and has been discussed in the context of speciation by, e.g., Mani and Clarke 1990, Barton 2001 and later authors. I think that the empirical evidence for fluctuating selection is also quite strong (e.g Bell 2010), and so it seems more important to discuss that, e.g., selection for robustness.
\end{point}

\reply{This is also a very interesting point worth discussing. We have added discussion on fluctuating selection \revref.
}

\begin{point}{Interpretation of the Haldane’s Rule result}
%4. Interpretation of the Haldane’s Rule result
%
  The abstract refers to ``another possible explanation'' of Haldane's Rule, and the explanation given is said to be quite different from the ``dominance theory'', because ``it derives from the nature of segregation variance''.  I wonder whether this is not misleading. My view is that the sort of model explored here (like additive quantitative trait model used by Barton 2001) naturally generates the sort of dominance relations required for the ``dominance theory'' to hold. Unless the authors strongly disagree, I would modify the abstract and \hyperlink{rev1:5}{lines 278-291}, or at least provide a fuller clarification for why the expanation is different from that discussed by Barton 2001.
\end{point}

\reply{
    Good point; on closer consideration we have decided to remove the reference to the dominance theory \revref,
    and have modified the introduction to say ``a concrete explanation'' rather than ``another explanation''.
    More generally, we have added a paragraph to the discussion about
    the connection between our (quantitative and essentially continuous) model and
    predictions based on the effects of single mutations on fitness. \llname{quantnote}
}

@@

\begin{point}{}
  \hyperlink{rev2:2}{(p. 2, l. 92)}: ``...more than one way to do the same thing...'' sounds a bit too colloquial. Consider changing it to ``...more than one way to accomplish the same outcome...'' or some other alternative phrasing.
\end{point}

\reply{Thank you for the suggestion, but we would prefer to keep the sentence as written. 
We understand your concern, but in order to make this paper easy to read,
we have avoided using acronyms and tried to use simple language wherever possible,
so long as the meaning remains clear.
}

\begin{point}{}
  \hyperlink{rev2:3}{(p. 5, l. 192)}: ``The Kalman decomposition then classifies each direction in kryptotype space as either reachable or unreachable''. Is there a biological motivation for considering unreachable directions in kryptotype space as degrees of freedom of the neutral space?
\end{point}

\reply{
    Good point; we've added some additional discussion of what the subspaces represent. \revref
}

@@

\begin{point}{}
\hyperlink{rev2:4}{(p. 6, l. 208)}: ``both the change of basis used to obtain the decomposition and, once in this form, all submatrices other than $A_{ro}$, $B_{ro}$, and $C_{ro}$ can be changed without affecting the phenotype, and so represent degrees of freedom''. A minor question on generality - does Kalman decomposition require that the coefficient matrices are time-invariant?
\end{point}

\reply{
  Yes, the Kalman decomposition is for linear time-invariant systems. We now note this \revref{}. 
}

\begin{point}{Recombination}
(p. 7) In the section ``Sexual reproduction and recombination'', we think that the recombination process is not clearly explained, making it more challenging to read the text from this point. We suggest that the authors clarify how exactly recombination is implemented in the model. Perhaps an example would help the reader to have a better picture of the process. This example could be inline, like the one introduced in the section on Haldane's rule.
\end{point}

\reply{
    Good point; this was unclear. We have clarified that we don't pick a particular model for recombination in this seciton. \revref
}

### 2021-08-10 kalman_walk_paper.tex
edits together

% It is not a new observation
The field of population genetics has also explored the consequences of the fact
%that phenotype does not prescribe a unique genotype
that there is often more than one way to do the same thing,
\hypertarget{rev2:2}
and observed that speciation might be the result of changes that are themselves neutral.
Indeed, \citet{bateson1909heredity} first proposed that what today we call a
Bateson-Dobzhansky-Muller incompatibility would likely arise through neutral changes. 
The potential for speciation has been analyzed in models of 
traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
in related fitness landscape models \citep{fraisse2016genetics},
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
% Recent work has shown that Fisher's geometric model of stabilizing selection
% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
It has also been shown that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. 
However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
and a regulatory cascade \citep{porter2002speciation}
found rapid speciation under directional selection
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
The rate at which hybrid incompatibility accumulates due to genetic drift creating
segregation variance between isolated populations is fairly well understood
\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche}, \revpoint{1}{2}
but model assumptions can strongly affect predictions,
including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
and the shape of the fitness landscape \citep{fraisse2016genetics}.
Our main aim is to provide a concrete framework 
that can provide natural predictions of these model parameters
across a general class of models.
Furthermore, tools from system theory allow analytical predictions to be made
for large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

This decomposition is unique up to a change of basis that preserves the block structure.
In particular, 
the minimal subsystem obtained by the Kalman decomposition
is unique up to a change of coordinates.
This implies that there is no equivalent system with a smaller number of kryptotypic dimensions
than the dimension of the minimal system.
It is remarkable that the gene regulatory network architecture to achieve a given input--output map is never unique --
both the change of basis used to obtain the decomposition\hypertarget{rev2:4}
and, once in this form, all submatrices other than $A_{\ro}$, $B_{\ro}$, and $C_{\ro}$ can be changed without affecting the phenotype,
and so represent degrees of freedom.
(However, some of these subspaces may affect how the system deals with noise.)

@@

First, we must specify how sexual reproduction acts on these systems.
Suppose that each of a diploid organisms' two genomes encodes a set of system coefficients
with the same kryptotype dimension. 
We assume that a diploid which has inherited systems $(A', B', C')$ and $(A'', B'', C'')$ from its two parents
has phenotype determined by the system that averages these two,
$((A'+A'')/2, (B'+B'')/2, (C'+C'')/2)$.

@@

\paragraph{Haldane's rule}\hypertarget{rev1:5}
This model naturally predicts
Haldane's rule, the observation that if only one hybrid sex is sterile or inviable 
it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex, orr1997haldane}.
For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
A male whose pair of haplotypes is 
$\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, 
  if dosage compensation upregulates heterogametes by a factor of two relative to homogametes (as with \emph{Drosophila}),
while a female homozygous for the haplotype
$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$, has phenotype determined by
$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
An $F_1$ male offspring of these two will have its phenotype determined by
$\left[\begin{smallmatrix} (A_1 + \bar A_1)/2 & (A_2 + \bar A_2)/2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
If both genes resided on the autosomes, this system would only be possible in an $F_2$ cross.
More generally, 
if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
$F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids, and recall that $F_2$s are significantly less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}).

%Although many alleles will be dominant if the phenotype--fitness relationship is convex,
%the underlying mechanism does not depend on the \emph{dominance theory} \citep{turelli1995dominance} to explain Haldane's rule:
%instead, it derives from the nature of segregation variance.
%
% In this setting, Haldane's rule is mechanistically similar to \emph{hybrid breakdown}, 
% the observation that $F_2$ hybrids will often be less fit than $F_1$ hybrid crosses, in the early stages of speciation.
% This mechanism relies on the genetic variation of a system being shared across the sex chromosomes and the autosomes,
% and appears to be consistent with the ``dominance'' theory \citep{orr1997haldane, coyne1998evolutionary, turelli1995dominance} which is often invoked to explain Haldane's rule.
% \plr{Explain that the effects on fitness are dominant.}
%and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
%\plr{what are they?}
%\citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$. In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule.

This model differs from the \emph{dominance theory} \citep{turelli1995dominance} in that it treats system coefficients as quantitative traits.
%The classification of alleles as either ``dominant'' or ``recessive'' fits most naturally for Mendellian traits. 
If we think of individual loci of large effect then the effect of mutations on fitness (if convex) will usually be recessive, congruent with the dominance theory. However, the dominance or recessiveness of even large effect mutations will depend on both the magnitude of their effect and precisely how they adjust system coefficients. 
%For such mutations that keep systems on the neutral manifold (imagine mutations to the oscillator that only move $\tau$), the dominance coefficient (usually notated as $h$) will be mathematically undefined. Mutations that driectly perturb coefficiens of $A$ can be either dominant or recessive, depending on their size.

\citet{barton2001role} explored Haldane's rule in the context of a quantitative trait and concluded that it was a consequence of the dominance theory. However, Barton also demonstrated $F_2$ hybrid breakdown in a model with strictly \emph{additive} inheritence -- a consequence of segregation variance. The model suggested here explains Haldane's rule as a special case of hybrid breakdown. As hybrid breakdown can occur for a quantitative trait under additivity, so can heterogametic inviability. This suggests that the model presented here is distinct from the dominance theory; like hybrid breakdown, Haldane's rule derives from the nature of segregation variance.

@@

Fluctuating selection \citep{bell2010fluctuating, barton2001role} could further accelerate speciation via system drift. 
Under our model, populations evolving towards different optimal phenotypes will always produce less fit hybrids,
as the hybrid phenotypes will be displaced from both parental optimums -- that is maladapted to both parental environments. This suggests that fluctuating selection -- if not synchronized between populations -- should diminish hybrid fitnesses.
In the case where fluctuations are synchronized, due to the size of the neutral system space, 
populations may arive at different system organizations when returning to previous phenotypes at a rate faster than by drift alone. \revpoint{1}{3}

@@

>>>>

% It is not a new observation
The field of population genetics has also explored the consequences of the fact
%that phenotype does not prescribe a unique genotype
that there is often more than one way to do the same thing, \revpoint{1}{2}
\hypertarget{rev2:2}
and observed that speciation might be the result of changes that are themselves neutral.
Indeed, \citet{bateson1909heredity} first proposed that what today we call a
Bateson-Dobzhansky-Muller incompatibility would likely arise through neutral changes. 
The potential for speciation has been analyzed in models of 
traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
in related fitness landscape models \citep{fraisse2016genetics},
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
% Recent work has shown that Fisher's geometric model of stabilizing selection
% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
It has also been shown that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. 
However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
and a regulatory cascade \citep{porter2002speciation}
found rapid speciation under directional selection
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
The rate at which hybrid incompatibility accumulates due to genetic drift creating
segregation variance between isolated populations is fairly well understood
\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
but model assumptions can strongly affect predictions,
including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
and the shape of the fitness landscape \citep{fraisse2016genetics}.
Our main aim is to provide a concrete framework 
that can provide natural predictions of these model parameters
across a general class of models.
Furthermore, tools from system theory allow analytical predictions to be made
for large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

This decomposition is unique up to a change of basis that preserves the block structure.
In particular, 
the minimal subsystem obtained by the Kalman decomposition
is unique up to a change of coordinates.
This implies that there is no equivalent system with a smaller number of kryptotypic dimensions
than the dimension of the minimal system.
It is remarkable that the gene regulatory network architecture to achieve a given input--output map is never unique --
both the change of basis used to obtain the decomposition\hypertarget{rev2:4}
and, once in this form, all submatrices other than $A_{\ro}$, $B_{\ro}$, and $C_{\ro}$ can be changed without affecting the phenotype,
and so represent degrees of freedom.

Although the other submatrices do not affect the phenotype
through the impulse response function $h(t)$,
they may affect individual fitness in other ways. \revpoint{2}{4}
For instance,
the ``unreachable'' subspaces cannot be affected by any of the possible variations
in input produced by varying $u(t)$.
However, if the nature of the input changed
(for instance, by perturbing a previously constant environmental variable),
differences in the unreachable components might lead to differences in phenotypes
and thus represent cryptic genetic variation.
Similarly, ``unobservable'' subspaces might represent aspects of phenotype
that are not constrained by selection,
but might become important if conditions change.
Additionally, all subspaces may affect how the system deals with noise.

@@

First, we must specify how sexual reproduction acts on these systems.
Suppose that each of a diploid organisms' two genomes encodes a set of system coefficients
with the same kryptotype dimension. 
We assume that a diploid which has inherited systems $(A', B', C')$ and $(A'', B'', C'')$ from its two parents
has phenotype determined by the system that averages these two,
$((A'+A'')/2, (B'+B'')/2, (C'+C'')/2)$.
(We do not, however, specify a model for recombination to produce the inherited systems,
as the details will depend on the genetic basis of the system coefficients;
two extremes are depicted in the two panels of Figure~\ref{fig:conceptual_fig}.) \revpoint{2}{6}

@@

\paragraph{Haldane's rule}\hypertarget{rev1:5}
This model naturally predicts
Haldane's rule, the observation that if only one hybrid sex is sterile or inviable 
it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex, orr1997haldane}.
For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
A male whose pair of haplotypes is 
$\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right]$, 
  if dosage compensation upregulates heterogametes by a factor of two relative to homogametes (as with \emph{Drosophila}),
while a female homozygous for the haplotype
$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$, has phenotype determined by
$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
An $F_1$ male offspring of these two will have its phenotype determined by
$\left[\begin{smallmatrix} (A_1 + \bar A_1)/2 & (A_2 + \bar A_2)/2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
If both genes resided on the autosomes, this system would only be possible in an $F_2$ cross.
More generally, 
if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
$F_1$ males are effectively equivalent to purely autosomal-system $F_2$ hybrids, and recall that $F_2$s are significantly less fit on average than $F_1$s (see Figure \ref{fig:osc_incompat}).
\revpoint{1}{4}

@@

Selection could dramatically accelerate speciation by essentially the same process --
for instance, \citet{mani1990mutational} showed that populations
developed incompatibilities as they independently evolved towards the same quantitative trait optimum,
as different alleles fix in different populations. \linelabel{citemani}
Fluctuating selection \citep{bell2010fluctuating, barton2001role} is ongoing, so could further accelerate speciation.
Under our model, populations evolving towards different optimal phenotypes will always produce less fit hybrids,
as the hybrid phenotypes will be displaced from both parental optimums -- that is maladapted to both parental environments. This suggests that fluctuating selection -- if not synchronized between populations -- should diminish hybrid fitnesses.
In the case where fluctuations are synchronized, due to the size of the neutral system space, 
populations may arive at different system organizations when returning to previous phenotypes at a rate faster than by drift alone. \revpoint{1}{3}

@@

\paragraph{Geometric models} \linelabel{quantnote}
Our results describe a continuum of possible systems,
described by their continuous coefficient matrices.
The arguments we make (e.g., Figure~\ref{fig:conceptual_fig}) are often geometric in nature,
and so it would be interesting to ask whether Fisher's geometric model
\citep[as in, e.g.,][]{martin2014,fraisse2016genetics,simon2017coadapted}
might provide more quantitative, generalizable predictions.
It is also natural to ask about the distribution of effect sizes of individual mutations
in a population evolving under this model,
and what the idstribituion of selection coefficients would be
(at equilibrium, perhaps).
These are questions that might be best answered by simulation,
and could provide much additional information -- for instance,
are alleles typically dominant in their effects on system coefficients?
or, fitness?
